---
title: "Satellite LiDAR Point Adjustment (salpa)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Satellite LiDAR Point Adjustment (salpa)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `salpa` package provides tools for optimizing and adjusting satellite LiDAR point positions using genetic algorithms. It focuses on two main types of adjustments:

1. Linear Alignment: Aligns points to their best-fit line using total least squares approach
2. Positional Correction: Fine-tunes point locations with specified offsets to minimize elevation differences by using genetic algorithm optimization

The package offers multiple distance metrics for comparing sequences, making it flexible for different types of analyses.

## Installation

You can install the development version of salpa from GitHub:

```r
# install.packages("remotes")
remotes::install_github("naru-T/salpa")
```

Note: The package is currently under development. Please check the GitHub repository at https://github.com/naru-T/salpa for the latest version and updates.

## Required Packages

First, let's load all necessary packages:

```{r setup, message=FALSE, warning=FALSE, eval=FALSE}
library(salpa)
library(sf)
library(terra)
library(exactextractr)
library(GA)
library(dtw)
library(tmap)
library(elevatr)
library(raster)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Data Preparation

We'll use a sample dataset of GEDI L2A shots (a series of footprints) from Nikko, Japan, included with the package. The data has been preprocessed using quality control flags (degrade_flag and quality_flag) to ensure data quality.  `elev_lowestmode` of the GEDI L2A data product has been considered as the GEDI-derived DEM. We'll compare these LiDAR footprints with a reference DEM:

```{r data_prep, eval=FALSE}
# Get the path to the sample data included with the package
gpkg_path <- system.file("extdata", "gedi_l2a_shot_sample.gpkg", package = "salpa")

# Read the LiDAR footprints
lidar_footprints <- st_read(gpkg_path)

# Define the area of interest
bbox_4326 <- st_bbox(c(xmin = 139.40119, ymin = 36.86163,
                       xmax = 139.52543, ymax = 36.94841),
                     crs = 4326)
names(bbox_4326) <- c("left", "bottom", "right", "top")
bbox_polygon <- st_as_sfc(bbox_4326)
bbox_sf <- st_as_sf(bbox_polygon)

```

## Reference DEM Preparation

We'll use a reference DEM to evaluate and optimize our adjustments:

```{r dem_prep, eval=FALSE}
# Get reference DEM
dem <- get_elev_raster(bbox_sf, z = 12, clip = "bbox")
dem_3857 <- projectRaster(dem, crs = 3857)
dem_rast <- rast(dem_3857)

# Extract DEM values for comparison
lidar_footprints_3857 <- st_transform(lidar_footprints, 3857)
original_dem <- exact_extract(dem_rast, lidar_footprints_3857 |> st_buffer(12.5), "mean", progress = FALSE)
```

## Positional Adjustment

we refine the point positions using genetic algorithm optimization. By default, the Euclidean distance metric is used, but you can specify other metrics based on your needs:

```{r positional_adjustment, eval=FALSE}
# Run positional correction optimization with default Euclidean distance
corrected_positions <- salpa::positional_correction(
  lidar_footprints = lidar_footprints_3857,
  input_rast = dem_rast,
  minimizing_method = "euclidean",  # Default method
  target_variable = "mean",
  lidar_value = "elev_lowestmode",
  buf = 12.5,
  crs_code = 3857,
  parallel = TRUE
)

# Extract DEM values for the final adjusted points
corrected_positions_dem <- exact_extract(dem_rast, corrected_positions[[2]] |> st_buffer(12.5), "mean", progress = FALSE)
```

`corrected_positions` is a list of sf objects. The first element is the original LiDAR footprints, and the second element is the adjusted points.
```{r print_corrected_positions, eval=FALSE}
print(corrected_positions[[1]])
print(head(corrected_positions[[2]]))
```


## Optional: Linear Alignment

The Linear alignment is optional. It aligns the LiDAR points to their best-fit line. This helps correct systematic linear offsets in the data:
Now This example applies linear alignment to the original LiDAR footprints.

```{r linear_alignment, eval=FALSE}
# Apply linear alignment
aligned_points <- linear_alignment(lidar_footprints, 3857)
linearly_adjusted_dem <- exact_extract(dem_rast, aligned_points |> st_buffer(12.5), "mean", progress = FALSE )
```

## Evaluation and Visualization

### 1. DEM Difference Trajectory

We can visualize how the adjustments affect the elevation differences:

```{r dem_difference, eval=FALSE}
# Create difference data frame
difference_df <- data.frame(
    index = 1:nrow(lidar_footprints),
    original = as.numeric(lidar_footprints$elev_lowestmode) -
              as.numeric(original_dem),
    linearly_adjusted = as.numeric(aligned_points$elev_lowestmode) -
                      as.numeric(linearly_adjusted_dem),
    positionally_corrected = as.numeric(corrected_positions[[2]]$elev_lowestmode) -
                      as.numeric(corrected_positions_dem)
)

# Convert to long format and plot
difference_long <- difference_df %>%
    tidyr::pivot_longer(cols = c(original, linearly_adjusted, positionally_corrected),
                names_to = "adjustment",
                values_to = "difference")

ggplot(difference_long, aes(x = index, y = difference, color = adjustment)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5, alpha = 0.7) +
    scale_color_manual(values = c("original" = "#2C3E50",
                                 "linearly_adjusted" = "#95A5A6",
                                 "positionally_corrected" = "#E74C3C"),
                      labels = c("Raw DEM",
                               "Linearly Aligned DEM",
                               "Positionally Adjusted DEM")) +
    labs(x = "Point Index",
         y = "DEM Difference (m)",
         color = "Adjustment Type") +
    theme_minimal()
```

### 2. Scatter Plot Comparison

Compare the relationships between LiDAR and DEM elevations:

```{r scatter_plots, eval=FALSE}
scatter_df <- data.frame(
    reference = as.numeric(lidar_footprints$elev_lowestmode),
    raw = as.numeric(original_dem),
    linear = as.numeric(linearly_adjusted_dem),
    location = as.numeric(corrected_positions_dem)
) %>%
    tidyr::pivot_longer(cols = c(raw, linear, location),
                names_to = "adjustment",
                values_to = "adjusted_value") %>%
    dplyr::mutate(adjustment = factor(adjustment,
                             levels = c("raw", "linear", "location")))

# Calculate R-squared values
r2_values <- scatter_df %>%
    group_by(adjustment) %>%
    summarise(r2 = round(cor(reference, adjusted_value)^2, 3)) %>%
    mutate(label = paste("R² =", r2))

ggplot(scatter_df, aes(x = adjusted_value, y = reference)) +
    geom_abline(intercept = 0, slope = 1, color = "grey50",
                linetype = "dashed", linewidth = 0.5) +
    geom_point(alpha = 0.6, size = 1.5, color = "#2C3E50") +
    geom_text(data = r2_values,
              aes(label = label),
              x = Inf, y = -Inf,
              hjust = 1.1, vjust = -1,
              size = 4) +
    facet_wrap(~adjustment,
               labeller = as_labeller(c(raw = "Raw DEM",
                                      linear = "Linearly Aligned DEM",
                                      location = "Positionally Adjusted DEM"))) +
    labs(y = "LiDAR Reference Elevation (m)",
         x = "Adjusted DEM Elevation (m)") +
    theme_minimal()
```

### 3. Mean Absolute Error (MAE) Analysis

Calculate MAE for each adjustment method:

```{r mae_calc, eval=FALSE}
mae_df <- data.frame(
    reference = as.numeric(lidar_footprints$elev_lowestmode),
    raw = as.numeric(original_dem),
    linear = as.numeric(linearly_adjusted_dem),
    location = as.numeric(corrected_positions_dem)
)

mae_values <- mae_df %>%
    pivot_longer(cols = c(raw, linear, location),
                names_to = "adjustment",
                values_to = "adjusted_value") %>%
    group_by(adjustment) %>%
    summarise(mae = mean(abs(reference - adjusted_value)))

print(mae_values)
```

### 4. Spatial Visualization

Finally, visualize the spatial distribution of points:

```{r spatial_viz, eval=FALSE}
tmap_mode("plot")
tm <- tm_shape(corrected_positions[[2]], bbox = bbox_4326) +
        tm_symbols(col = "red", shape = 20, size = 0.1) +
    tm_shape(lidar_footprints, bbox = bbox_4326) +
        tm_symbols(col = "black", shape = 3, size = 0.1) +
    tm_shape(linearly_adjusted_sf, bbox = bbox_4326) +
        tm_symbols(col = "#b6b4b4", shape = 4, size = 0.1) +
    tm_layout(legend.position = c("left", "top")) +
    tm_add_legend(type = "symbols",
                  labels = c("Positionally Adjusted",
                           "Original",
                           "Linearly Aligned"),
                  shape = c(20, 3, 4),
                  fill = c("red", "black", "#b6b4b4"),
                  size = c(1, 1, 1.1),
                  title = "Point Types")
tm
```

## Conclusion

The `salpa` package provides a comprehensive workflow for adjusting satellite LiDAR point positions:

1. Positional adjustment refines point locations using genetic algorithm optimization
2. Linear alignment is optional. Itcorrects systematic offsets by aligning points to their best-fit line. However, in this demo, the result is not well presented.
3. The positional adjustment typically results in better alignment with reference DEMs, as shown by:
   - Reduced elevation differences
   - Improved R² values in scatter plots
   - Lower Mean Absolute Error (MAE)

This approach helps improve the spatial accuracy of satellite LiDAR data, making it more suitable for various applications in satellite LiDAR data analysis.

## Distance Metrics

The package provides several distance metrics for comparing sequences:

1. **Euclidean Distance** (Default)
   - Standard point-to-point distance calculation
   - Computed only on complete pairs
   - Normalized by the number of complete pairs
   - Best for direct value comparisons

2. **Dynamic Time Warping (DTW)**
   - Handles temporal shifts and warping
   - Uses step patterns that can skip NA values
   - Useful for sequences with timing differences

3. **Manhattan Distance**
   - Sum of absolute differences
   - Less sensitive to outliers than Euclidean
   - Good for grid-like movements

4. **Correlation Distance**
   - Based on Pearson correlation
   - Captures pattern similarity regardless of scale
   - Requires at least 2 complete pairs

5. **Area Distance**
   - Measures area between curves
   - Uses linear interpolation for NA values
   - Good for overall trajectory comparison

6. **Hausdorff Distance**
   - Maximum minimum distance between sequences
   - Captures worst-case deviations
   - Considers only non-NA values

Each metric handles NA values appropriately and can be selected using the `minimizing_method` parameter in relevant functions.
