---
title: "Satellite LiDAR Point Adjustment (slpa)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Satellite LiDAR Point Adjustment (slpa)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `slpa` package provides tools for optimizing and adjusting satellite LiDAR point positions using genetic algorithms. It focuses on linear adjustment and location correction of spatial point data, particularly for satellite-based LiDAR observations.

## Installation

You can install the development version of slpa from GitHub with:

```r
# install.packages("devtools")
devtools::install_github("yourusername/slpa")
```

## Basic Usage

First, load the package:

```{r setup}
library(slpa)
```

The package provides several key functions for point adjustment:

1. `linear_adjust()`: Aligns points to a best fit line using genetic algorithm optimization
2. `location_correction()`: Corrects point locations with specified offsets

For detailed examples and usage, please refer to the function documentation.
